---
title: "ASSIGNMENT 1: ADVANCED R PROGRAMMING"
author: "MUSKAAN ARORA: 23205389"
format:
  revealjs:
    smaller: true
    scrollable: true
    theme: moon 
output:
  revealjs:
    center: true             
    slideNumber: true        
    controls: true           
    progress: true           
    history: true            
    transition: slide        
    width: 1200              
    height: 800              
    incremental: true        
    highlight: tango         
    code_folding: show       
---

## 1. Reading the data

-   **Objective**: Loading and preprocessing two datasets (`hdro_indicators_ind.csv` and `hdro_indicators_irl.csv`) using the `data.table` package.

    -   Removing the first row from each dataset (`hdro_ind` and `hdro_irl`) assuming it contains headers or metadata.

```{r, echo = TRUE}
# Installing and loading data.table 
#install.packages("data.table")
library(data.table)

# Loading the data into data.table
hdro_ind <- fread("hdro_indicators_ind.csv")
hdro_irl <- fread("hdro_indicators_irl.csv")

# Remove the first row from hdro_ind
hdro_ind <- hdro_ind[-1,]

# Remove the first row from hdro_irl
hdro_irl <- hdro_irl[-1,]
```

## 2. Merging the Datasets

-   **Objective**: Combining the two datasets (`hdro_ind` and `hdro_irl`) into one (`merged_hdro`) using the rbind fuction

    -   Displaying the first few rows (`head`) of the merged dataset to inspect the structure and ensure the merge was successful.

```{r, echo=TRUE}
# Merging the datasets 
merged_hdro <- rbind(hdro_ind, hdro_irl)

# View the merged data
head(merged_hdro)
```

## 3. Data exploration

-   **Objective**: Perform basic data exploration and preprocessing on `merged_hdro`.

    -   Converting the `value` column to numeric.

    -   Computing a summary of the dataset to understand its distribution and characteristics.

    -   Calculating quantiles of the value column to identify key statistical measures (25th, 50th, and 75th percentiles).

```{r, echo=TRUE}
#converting value to numeric
merged_hdro[, value := as.numeric(value)]

#summary of the merged data
summary(merged_hdro)

# Quantiles of 'value'
quantiles_value <- quantile(merged_hdro$value, probs = c(0.25, 0.5, 0.75),na.rm=TRUE)
print(quantiles_value)
```

## 4. Analysis using data.table

-   **Objective**: Perform deeper analysis using `data.table` functionalities on `merged_hdro`.

    -   Calculate summary statistics (`mean`, `median`, `min`, `max`, `sd`) by `country_name` and `year` using `data.table` syntax (`keyby`).

```{r,echo=TRUE}
# Convert merged_data to data.table
setDT(merged_hdro)

#Summary statistics by 'country_name'
summary_by_country <- merged_hdro[, .(mean_value = mean(value, na.rm = TRUE),
                                      median_value = median(value, na.rm = TRUE),
                                      min_value = min(value, na.rm = TRUE),
                                      max_value = max(value, na.rm = TRUE),
                                      sd_value = sd(value, na.rm = TRUE)),
                                  keyby = country_name]

print(summary_by_country)

#Summary statistics by 'year' and 'country'
summary_by_year_country <- merged_hdro[, .(mean_value = mean(value, na.rm = TRUE),
                                      median_value = median(value, na.rm = TRUE),
                                      min_value = min(value, na.rm = TRUE),
                                      max_value = max(value, na.rm = TRUE),
                                      sd_value = sd(value, na.rm = TRUE)),
                                  keyby = .(year,country_name)]

print(summary_by_year_country)
```

## 5. Plotting

-   **Objective**: Visualizing insights derived from the data using `ggplot2`.

    -   Create a bar plot showing the mean value by `country_name`.

    -   Generate a line plot illustrating the mean value trends over `year` for each `country_name`.

```{r,echo=TRUE}
library(ggplot2)

# Plot of mean value by country_name
ggplot(summary_by_country, aes(x = reorder(country_name, mean_value), y = mean_value)) +
  geom_bar(stat = "identity", fill = "blue", color = "black") +
  labs(title = "Mean Value by Country",
       x = "Country",
       y = "Mean Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Plot of mean value by year and country
ggplot(summary_by_year_country, aes(x = year, y = mean_value, group = country_name, color = country_name)) +
  geom_line() +
  labs(title = "Mean Value by Year and Country",
       x = "Year",
       y = "Mean Value",
       color = "Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```
